---
title: "EDS 213 Final Project"
subtitle: "National Park Popularity Query"
author: "Nicole Pepper"
date: "2025-05-14"
format: html
editor_options: 
  chunk_output_type: console
---

# My Query:
For my query I will be using my database on National Park Popularity Database to explore which parks are the most popular with both humans and wildlife.

# My Data:

- NPS Species Occurrence from NPSpecies: https://github.com/frankiethull/NPSpecies
- NPS Visitation Data from NPS Visitor Use Statistics Dashboard: https://www.nps.gov/subjects/socialscience/visitor-use-statistics-dashboard.htm

# My Questions:

- Which parks are favored by humans?
- Which parks are favored by wildlife? 
- Do humans and wildlife have any favorites in common?

Set Up Workspace
```{r}
library(duckdb)
library(dbplyr)
library(DBI)
library(tidyverse)
library(scales)
library(ggbump)
```

Read in Data
```{r}

file.copy(
  from = "data/clean/national_parks.duckdb",
  to   = "data/clean/national_parks_analysis.duckdb",
  overwrite = TRUE)

conn <- dbConnect(duckdb::duckdb(), dbdir = "data/clean/national_parks_analysis.duckdb", read_only = FALSE)
```

## Queries:

### 1)  Identify Top 10 Most Popular Parks for Humans
```{sql}

-- Identify Top 10 Most Popular Parks for Humans

top_human <- dbGetQuery(conn, "
-- select fields of interest
SELECT 
  ID,
  FullName,
  Region,
  State,
  -- sum total visits
  SUM(RecreationVisits) AS TotalVisits
FROM NP_Visits
-- join with NP key using ID
JOIN NP_Key USING (ID)
GROUP BY ID, FullName, Region, State
-- select top 10
ORDER BY TotalVisits DESC
LIMIT 10;
")



```

Plot Top 10 Most Popular Parks for Humans
```{r}
# Plot  Top 10 Most Popular Parks for Humans
top_human_plot <- top_human |>
  mutate(ParkLabel = paste0(FullName, ", ", State))|>
  ggplot(aes(x = TotalVisits, y = reorder(ParkLabel, TotalVisits))) +
  geom_col(fill = "orange") +
  geom_text(aes(label = label_number(scale = 1e-9,
                                     suffix = " M",
                                     accuracy = 0.01)(TotalVisits)),
            hjust = -0.1,
            size = 5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(title = "Human's Favorite National Parks",
    x = "Total Visits (2024)",
    y = "National Park") +
  theme_minimal(base_size = 20) +
  theme(plot.title.position = "plot",
        panel.grid.major.y = element_blank(),
        axis.text.x = element_blank())

top_human_plot

```



## 2)  Identify Top 10 Most Popular Parks for Wildlife by Total Species Count
```{sql}
-- Identify Top 10 Most Popular Parks for Wildlife by Total Species Count

top_wildlife <- dbGetQuery(conn, "
-- select fields of interest
SELECT 
  NP_Key.ID,
  NP_Key.FullName,
  NP_Key.Region,
  NP_Key.State,
  -- sum total observations
  SUM(NP_Species_History.Observations) AS TotalAnimals
FROM NP_Species_History
-- join with NP key using ID
JOIN NP_Key ON NP_Species_History.ParkCode = NP_Key.ID
GROUP BY NP_Key.ID, NP_Key.FullName, NP_Key.Region, NP_Key.State
-- select top 10
ORDER BY TotalAnimals DESC
LIMIT 10;
")


```

Plot Top 10 Most Popular Parks for Wildlife by Total Species Count
```{r}
#  Plot Top 10 Most Popular Parks for Wildlife by Total Species Count
top_wildlife_plot <- top_wildlife |>
  mutate(
    ParkLabel = paste0(FullName, ", ", State),
    LabelFormatted = case_when(
      TotalAnimals >= 1e6 ~ paste0(round(TotalAnimals / 1e6, 2), " M"),
      TotalAnimals >= 1e3 ~ paste0(round(TotalAnimals / 1e3), " K"),
      TRUE ~ as.character(TotalAnimals))) |>
  ggplot(aes(x = TotalAnimals, y = reorder(ParkLabel, TotalAnimals))) +
  geom_col(fill = "orange") +
  geom_text(aes(label = LabelFormatted),
            hjust = -0.1,
            size = 5) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Wildlife's Favorite National Parks - Total Wildlife",
    x = "Total Wildlife Counts (2024)",
    y = "National Park") +
  
  theme_minimal(base_size = 20) +
  
  theme(
    plot.title.position = "plot",
    panel.grid.major.y = element_blank(),
    axis.text.x = element_blank())


top_wildlife_plot
```


3)  Identify Top 10 Most Popular Parks for Wildlife by Unique Species Count
```{sql}
-- Identify Top 10 Most Popular Parks for Wildlife by Unique Species Count
top_unique_species <- dbGetQuery(conn, "
-- select fields of interest
SELECT 
  NP_Key.ID,
  NP_Key.FullName,
  NP_Key.Region,
  NP_Key.State,
    -- count total unique (distinct) species
  COUNT(DISTINCT NP_Species_History.TaxonCode) AS SpeciesCount
FROM
  NP_Species_History
  -- join with NP key using ID
JOIN NP_Key ON NP_Species_History.ParkCode = NP_Key.ID
GROUP BY NP_Key.ID, NP_Key.FullName, NP_Key.Region, NP_Key.State
-- select top 10
ORDER BY SpeciesCount DESC
LIMIT 10;")

```

Plot Top 10 Most Popular Parks for Wildlife by Unique Species Count
```{r}
# Plot Top 10 Most Popular Parks for Wildlife by Unique Species Count
top_unique_species_plot <- top_unique_species |>
  mutate(ParkLabel = paste0(FullName, ", ", State))|>
  ggplot(aes(x = SpeciesCount, y = reorder(ParkLabel, SpeciesCount))) +
  geom_col(fill = "orange") +
  geom_text(aes(label = label_comma()(SpeciesCount)),
            hjust = -0.1,
            size = 5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(
    title = "Wildlife's Favorite National Parks - Most Unique Species",
    x = "Total Unique Species (2024)",
    y = "National Park") +
  theme_minimal(base_size = 20) +
  theme(plot.title.position = "plot",
        panel.grid.major.y = element_blank(),
        axis.text.x = element_blank())

top_unique_species_plot

```

```{r}

# --- Find where humans and wildlife have favorites in common ----

# Get unique park names for each dataset
park_names <- top_human$FullName
total_wildlife_names <- top_wildlife$FullName
unique_wildlife_names <- top_unique_species$FullName

# Find where they match
in_all_three <- list(Reduce(intersect, list(park_names, total_wildlife_names, unique_wildlife_names)))
human_and_total <- list(intersect(park_names, total_wildlife_names))
human_and_unique <- list(intersect(park_names, unique_wildlife_names))

```
